$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-06 paper/paper.tex
Forgotten todo
Fix github link
Introduce some suggested changes

\usepackage[dvipsnames]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{todonotes}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage{amsbsy}
\usepackage{mathbbol}
\usepackage{mathtools}
\usepackage[many]{tcolorbox}
\bibliographystyle{plainurl}

@@

% TODO notes -- TODO: remove
\usepackage{todonotes}

@@

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\; \type{\longrightarrow}_{#1} \;}
\newcommand{\types}[4]{#1 \; \type{;} \; #2 \; \type{\vdash} \; #3 \; \type{\triangleright} \; #4}
\newcommand{\contains}[6]{#1 \; \type{;} \; #2 \; \type{\ni}_{#3} \; #4 \; \type{;} \; #5 \; \type{\triangleright} \; #6}
\newcommand{\containsusage}[4]{#1 \; \type{\ni}_{#2} \; #3 \; \type{\triangleright} \; #4}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\;}
\newcommand{\Idx}{\type{IDX}\;}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Usage}{\type{USAGE}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Algebra}{\type{ALGEBRA}}
\newcommand{\eq}[1]{\; \type{=}_{#1} \;}
\newcommand{\eqeq}{\; \type{\equiv} \;}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{$\pi$ with leftovers: \\ a mechanisation in Agda}
\titlerunning{$\pi$ with leftovers: all mechanisation in Agda}
\author{Uma Zalakain}{University of Glasgow, Scotland}
       {u.zalakain.1@research.gla.ac.uk}{https://orcid.org/0000-0002-3268-9338}{}
\author{Ornela Dardha}{University of Glasgow, Scotland}
       {ornela.dardha@glasgow.ac.uk}{https://orcid.org/0000-0001-9927-7875}{}
\authorrunning{U. Zalakain and O. Dardha}
\Copyright{Uma Zalakain and Ornela Dardha}
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10003761.10003764</concept_id>
<concept_desc>Theory of computation~Process calculi</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[300]{Theory of computation~Process calculi}
\keywords{pi-calculus, linear types, leftover typing, concurrency, mechanisation, Agda}
\supplement{\url{https://github/umazalakain/typing-linear-pi}}
\acknowledgements{We want to thank Erika Kreuter, Wen Kokke, James Wood, Guillaume Allais, Bob Atkey, and Conor McBride for their thoughts, patience, work, education and camaraderie.}
\funding{Supported by the UK EPSRC grant EP/K034413/1, ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the EU HORIZON 2020 MSCA RISE project 778233
  ``Behavioural Application Program Interfaces'' (BehAPI).}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{nidefinition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$, that is \emph{partial}, \emph{functional}, \emph{cancellative}, \emph{associative}, and \emph{commutative}.
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal}.
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} uses a terniary relation to model such an algebra as a record $\Algebra_C$ on a carrier $C$.

@@

  $\func{\_}$ : $\types{\constr{[]} \comma \unit}{\constr{[]} \comma \func{\omega}}{\func{p}}{\func{epsilon}}$
  $\func{\_}$ = $\constr{chan} \; \channel{\unit}{\func{\omega}} \; \li \; \one \; (\constr{comp}$
       $(\constr{recv} \; \constr{0} \; (\constr{recv} \; \constr{0} \; \constr{end}))$
       $(\constr{chan} \; \unit \; \func{\omega} \; \one \; (\constr{send} \; (\suc \; \constr{0}) \; \constr{0} \; (\constr{send} \; \constr{0} \; (\suc \; \suc \; \constr{0}) \; \constr{end})))
       )$
  \end{lstlisting}
\end{example}

@@

\subparagraph*{Substitution}
\todo{intro}
\autoref{thm:subst-generalization} proves a generalised version of substitution, where the substitution $\subst{P}{j}{i}$ is on any variable $i$.
\autoref{thm:substitution} instantiates the generalised version to the concrete case where $i$ is the most recently introduced variable $\constr{0}$, as required by subject reduction.

@@

Allais \cite{Allais2018a} uses leftover typing to mechanise in Agda a bidirectional type system for the linear \lambdacalc{}.
He proves type preservation and provides a decision procedure for type checking and type inference.
In our paper, we apply leftover typing from Allais \cite{Allais2018a} to the \picalc{} and focus on type checking. We give a more general usage algebra, leading to linear, graded and shared type systems.
Drawing from quantitative type theory (by McBride and Atkey \cite{McBride2016, Atkey2018}), in our work we are also able to talk about fully consumed resources.
As such, we can transmit over a channel all none multiplicities of a fully exhausted channel, namely a channel of multiplicity $\lz$.

>>>>

\usepackage[dvipsnames]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage{amsbsy}
\usepackage{mathbbol}
\usepackage{mathtools}
\usepackage[many]{tcolorbox}
\bibliographystyle{plainurl}

@@

@@

% Types
\newcommand{\Set}{\type{SET}}
\newcommand{\reduce}[1]{\; \type{\longrightarrow}_{#1} \;}
\newcommand{\types}[4]{#1 \; \type{;} \; #2 \; \type{\vdash} \; #3 \; \type{\triangleright} \; #4}
\newcommand{\contains}[6]{#1 \; \type{;} \; #2 \; \type{\ni}_{#3} \; #4 \; \type{;} \; #5 \; \type{\triangleright} \; #6}
\newcommand{\containsusage}[4]{#1 \; \type{\ni}_{#2} \; #3 \; \type{\triangleright} \; #4}
\newcommand{\Var}{\type{VAR}}
\newcommand{\Process}{\type{PROCESS}}
\newcommand{\Unused}{\type{UNUSED}}
\newcommand{\PreCtx}{\type{PRECTX}}
\newcommand{\Ctx}{\type{CTX}}
\newcommand{\Type}{\type{TYPE}\;}
\newcommand{\Idx}{\type{IDX}\;}
\newcommand{\Idxs}{\type{IDXS}}
\newcommand{\Usage}{\type{USAGE}}
\newcommand{\N}{\type{\mathbb{N}}}
\newcommand{\Channel}{\type{CHANNEL}}
\newcommand{\Rec}{\type{REC}}
\newcommand{\Algebra}{\type{ALGEBRA}}
\newcommand{\eq}[1]{\; \type{=}_{#1} \;}
\newcommand{\eqeq}{\; \type{\equiv} \;}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{$\pi$ with leftovers: \\ a mechanisation in Agda}
\titlerunning{$\pi$ with leftovers: all mechanisation in Agda}
\author{Uma Zalakain}{University of Glasgow, Scotland}
       {u.zalakain.1@research.gla.ac.uk}{https://orcid.org/0000-0002-3268-9338}{}
\author{Ornela Dardha}{University of Glasgow, Scotland}
       {ornela.dardha@glasgow.ac.uk}{https://orcid.org/0000-0001-9927-7875}{}
\authorrunning{U. Zalakain and O. Dardha}
\Copyright{Uma Zalakain and Ornela Dardha}
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10003761.10003764</concept_id>
<concept_desc>Theory of computation~Process calculi</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[300]{Theory of computation~Process calculi}
\keywords{pi-calculus, linear types, leftover typing, concurrency, mechanisation, Agda}
\supplement{\url{https://github.com/umazalakain/typing-linear-pi}}
\acknowledgements{We want to thank Erika Kreuter, Wen Kokke, James Wood, Guillaume Allais, Bob Atkey, and Conor McBride for their thoughts, patience, work, education and camaraderie.}
\funding{Supported by the UK EPSRC grant EP/K034413/1, ``From Data Types to Session Types: A Basis for Concurrency and Distribution'' (ABCD), and by the EU HORIZON 2020 MSCA RISE project 778233
  ``Behavioural Application Program Interfaces'' (BehAPI).}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{nidefinition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$, that is \emph{partial}, \emph{functional}, \emph{cancellative}, \emph{associative}, and \emph{commutative}.
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal}.
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} uses a ternary relation to model such an algebra as a record $\Algebra_C$ on a carrier $C$.

@@

  $\func{\_}$ : $\types{\constr{[]} \comma \unit}{\constr{[]} \comma \func{\omega}}{\func{p}}{\func{\epsilon}}$
  $\func{\_}$ = $\constr{chan} \; \channel{\unit}{\func{\omega}} \; \li \; \one \; (\constr{comp}$
       $(\constr{recv} \; \constr{0} \; (\constr{recv} \; \constr{0} \; \constr{end}))$
       $(\constr{chan} \; \unit \; \func{\omega} \; \one \; (\constr{send} \; (\suc \; \constr{0}) \; \constr{0} \; (\constr{send} \; \constr{0} \; (\suc \; \suc \; \constr{0}) \; \constr{end})))
       )$
  \end{lstlisting}
\end{example}

@@

\subparagraph*{Substitution}
\autoref{thm:subst-generalization} proves a generalised version of substitution, where the substitution $\subst{P}{j}{i}$ is on any variable $i$.
\autoref{thm:substitution} instantiates the generalised version to the concrete case where $i$ is the most recently introduced variable $\constr{0}$, as required by subject reduction.

@@

Allais \cite{Allais2018a} uses leftover typing to mechanise in Agda a bidirectional type system for the linear \lambdacalc{}.
He proves type preservation and provides a decision procedure for type checking and type inference.
In our paper, we apply leftover typing from Allais \cite{Allais2018a} to the \picalc{} and focus on type checking. We give a more general usage algebra, leading to linear, graded and shared type systems.
Drawing from quantitative type theory (by McBride and Atkey \cite{McBride2016, Atkey2018}), in our work we are also able to talk about fully consumed resources.
As such, we can transmit over a channel $\lz$ multiplicities of a fully exhausted channel.
